---
title: "Longitudinal and Survival Analysis of a Simulated Clinical Follow-up Study"
subtitle: "Biomarker trajectories and time to death under standard vs new treatment"
author: "Tanvir Ahmed"
date: "`r format(Sys.Date(), '%B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: true
    code_folding: show
    df_print: kable
    theme: flatly
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 8, fig.height = 4.8, dpi = 150, fig.align = "center"
)
```

> **About this project.** This analysis was originally completed as the final
> assignment of a university course in *Longitudinal and Survival Data Analysis*.
> The dataset is simulated and was provided by the course instructor. The
> analysis has since been rewritten as a single reproducible document, and the
> statistical reasoning has been revised (see [What was revised](#revisions)).

# Study and research questions

Data come from a simulated clinical follow-up study of **240 adults with a
chronic disease**, allocated to **standard treatment** (`treatment = 0`, n = 134)
or a **new treatment** (`treatment = 1`, n = 106).

* A continuous **biomarker** was measured at baseline and at 3, 6 and 12 months.
* **Survival** (time to death, in years) was then followed for up to seven years.

| Variable | Description |
|---|---|
| `id` | patient identifier |
| `treatment` | 0 = standard, 1 = new treatment |
| `age` | age at baseline (years) |
| `sex` | 0 = female, 1 = male |
| `bio_0`, `bio_3`, `bio_6`, `bio_12` | biomarker at 0, 3, 6, 12 months |
| `follow` | survival follow-up time (years) |
| `death` | 1 = died, 0 = censored |

The analysis answers four questions:

1. **How does the biomarker change over time, and does the change differ between treatments?**
   (exploratory plots, repeated-measures ANOVA, linear mixed models, GEE)
2. **How much of the biomarker variation is between patients rather than within them?**
   (variance components, ICC, random slopes)
3. **Does the new treatment improve survival?** (Kaplan–Meier, log-rank, Cox regression)
4. **Is the baseline biomarker prognostic, and does it change the estimated treatment effect?**

```{r packages}
library(tidyverse)   # data handling and plotting
library(afex)        # repeated-measures ANOVA
library(emmeans)     # estimated marginal means
library(lme4)        # linear mixed models
library(lmerTest)    # p-values for mixed-model fixed effects
library(geepack)     # generalized estimating equations
library(survival)    # Kaplan-Meier and Cox models
library(broom)
library(broom.mixed)
library(knitr)

theme_set(theme_minimal(base_size = 12))
trt_cols   <- c("Standard" = "#2C7FB8", "New" = "#D95F0E")
dir.create("figures", showWarnings = FALSE)
save_fig <- function(p, name, w = 8, h = 4.8) {
  ggsave(file.path("figures", name), p, width = w, height = h, dpi = 150, bg = "white")
  p
}
```

# Data preparation

```{r read}
data_wide <- read.table("data/extra_data.txt", header = TRUE) |>
  mutate(trt = factor(treatment, levels = 0:1, labels = c("Standard", "New")))

dim(data_wide)
anyNA(data_wide)
```

The data have one row per patient (240 × 10), no missing values, and the four
biomarker visits stored in separate columns (wide format). Longitudinal models
need one row per measurement occasion, so the data are reshaped to long format
with both a **numeric** time (months; for modelling trends) and a **categorical**
time (for visit-by-visit comparisons without assuming a shape).

```{r long}
data_long <- data_wide |>
  pivot_longer(starts_with("bio_"), names_to = "visit", values_to = "biomarker") |>
  mutate(
    time_numeric = as.numeric(sub("bio_", "", visit)),
    time_factor  = factor(time_numeric, levels = c(0, 3, 6, 12)),
    id           = factor(id),
    age_c        = age - mean(data_wide$age)   # centred age, used in the mixed models
  )

table(data_long$time_factor)
stopifnot(nrow(data_long) == 240 * 4)
```

Each of the four visits contains exactly 240 measurements (240 × 4 = 960 rows),
so the restructuring is complete and every patient contributes a full profile.

**Baseline characteristics by group**

```{r baseline}
data_wide |>
  group_by(Treatment = trt) |>
  summarise(
    n            = n(),
    `Age, mean (SD)`        = sprintf("%.1f (%.1f)", mean(age), sd(age)),
    `Male, %`               = sprintf("%.0f", 100 * mean(sex)),
    `Baseline biomarker, mean (SD)` = sprintf("%.1f (%.1f)", mean(bio_0), sd(bio_0)),
    Deaths       = sum(death)
  ) |>
  kable()

t.test(bio_0 ~ trt, data = data_wide)
```

An important observation for later: the groups **already differ at baseline**
in the biomarker (about 3 units lower in the new-treatment group, p < 0.001).
In a randomised trial this would be a chance imbalance; either way it matters
when interpreting treatment "main effects" and when adjusting for `bio_0` in the
survival models.

# Exploratory analysis

```{r fig-trajectories, fig.height = 4.5}
mean_profiles <- data_long |>
  group_by(trt, time_numeric) |>
  summarise(mean = mean(biomarker), se = sd(biomarker) / sqrt(n()), .groups = "drop")

p_traj <- ggplot(data_long, aes(time_numeric, biomarker)) +
  geom_line(aes(group = id, colour = trt), alpha = 0.18, linewidth = 0.35) +
  geom_line(data = mean_profiles, aes(y = mean, colour = trt), linewidth = 1.4) +
  geom_point(data = mean_profiles, aes(y = mean, colour = trt), size = 2.6) +
  scale_colour_manual(values = trt_cols) +
  scale_x_continuous(breaks = c(0, 3, 6, 12)) +
  labs(x = "Months since baseline", y = "Biomarker", colour = "Treatment",
       title = "Individual biomarker trajectories with group means") +
  theme(legend.position = "top")

save_fig(p_traj, "01_trajectories.png")
```

```{r mean-table}
mean_profiles |>
  select(-se) |>
  pivot_wider(names_from = time_numeric, values_from = mean, names_prefix = "Month ") |>
  mutate(`Change 0→12` = `Month 12` - `Month 0`) |>
  kable(digits = 1)
```

**What the plot shows**

* **Between-subject variability is large.** Baseline values range from about 46
  to 80, and patients keep their relative position over time (lines run roughly
  parallel), which points to a strong patient-level random intercept.
* **Within-subject trend.** Almost all trajectories decline, roughly linearly in
  time, with visit-to-visit noise around each patient's line.
* **Treatment difference.** The new-treatment group falls by about 17.5 units over
  12 months compared with about 10.6 units under standard treatment, so the group
  means diverge over time. The new group also starts about 3 units lower.

These are descriptive impressions; the models below test them formally.

# Repeated-measures ANOVA

Treatment is a between-subject factor and visit a within-subject factor. The
classical repeated-measures ANOVA assumes **sphericity** (implied by compound
symmetry: equal variances at all visits and equal correlation between any two
visits).

```{r rmanova}
anova_rm <- aov_car(
  biomarker ~ trt * time_factor + Error(id / time_factor),
  data = data_long,
  anova_table = list(correction = "none", es = "pes")
)
anova_rm
```

```{r rmanova-emm}
emmeans(anova_rm, ~ trt | time_factor) |> as_tibble() |>
  select(time_factor, trt, emmean, SE) |> kable(digits = 2)
```

* **Time** (F = 721.9, p < 0.001): the biomarker changes strongly over follow-up.
* **Treatment** (F = 129.2, p < 0.001): averaged over visits, the new-treatment
  group is lower. Part of this is the baseline imbalance noted above, so the main
  effect alone should not be read as a treatment effect.
* **Treatment × time** (F = 42.5, p < 0.001): the groups change at different
  rates. This interaction is the effect of interest: the gap grows from about
  3 units at baseline to about 10 units at 12 months.

## Checking the assumptions

```{r sphericity}
summary(anova_rm)$sphericity.tests
summary(anova_rm)$pval.adjustments
```

```{r gg-corrected}
nice(anova_rm, correction = "GG")
```

```{r cov-structure}
wide_bio <- data_wide |> select(bio_0, bio_3, bio_6, bio_12)
round(apply(wide_bio, 2, var), 1)       # variances by visit
round(cor(wide_bio), 2)                 # correlations between visits
```

```{r fig-rm-resid, fig.height = 3.6}
rm_resid <- tibble(resid = residuals(anova_rm))

p_qq <- ggplot(rm_resid, aes(sample = resid)) +
  stat_qq(alpha = 0.4, size = 0.9) + stat_qq_line(colour = "#D95F0E") +
  labs(title = "Normal Q-Q plot of RM-ANOVA residuals",
       x = "Theoretical quantiles", y = "Residual")
save_fig(p_qq, "02_rmanova_qq.png", h = 3.6)
```

**Assessment.**

* **Normality:** residuals from the repeated-measures model lie close to the
  Q-Q line, with only slight deviation in the tails.
* **Sphericity:** this is the condition the within-subject F-tests actually
  require. It holds: Mauchly's test is far from significant (p = 0.99) and the
  Greenhouse–Geisser ε ≈ 1.00, so the corrected tests are identical to the
  uncorrected ones.
* **Compound symmetry** is a stronger, sufficient condition, and it does *not*
  strictly hold. The pooled variance grows from about 30 at baseline to about
  55 at 12 months, partly because the two groups drift apart. The between-visit
  correlations are similar (0.60–0.75) but not equal. Sphericity can hold
  without compound symmetry, so the F-tests remain valid.
* **Other limits:** RM-ANOVA treats time as unordered categories, cannot adjust
  for continuous covariates such as age, and drops any patient with a missing
  visit. The mixed models below address all three.

# Linear mixed model with a random intercept

$$
\text{bio}_{ij} = \beta_0 + \beta_1 t_{ij} + \beta_2\,\text{trt}_i + \beta_3\,(t_{ij}\times\text{trt}_i)
+ \beta_4\,\text{age}^{c}_i + \beta_5\,\text{sex}_i + b_{0i} + \varepsilon_{ij},
\qquad b_{0i}\sim N(0,\sigma_b^2),\ \varepsilon_{ij}\sim N(0,\sigma^2)
$$

Age is **centred at the sample mean** (`r round(mean(data_wide$age), 1)` years),
so the intercept is the expected baseline value for a female patient of average
age under standard treatment, rather than a value at age 0.

```{r lmm-ri}
model_ri <- lmer(
  biomarker ~ time_numeric * treatment + age_c + sex + (1 | id),
  data = data_long, REML = TRUE
)

fixef_tab <- tidy(model_ri, effects = "fixed", conf.int = TRUE) |>
  select(term, estimate, std.error, conf.low, conf.high, p.value)
kable(fixef_tab, digits = c(0, 3, 3, 3, 3, 4),
      col.names = c("Term", "Estimate", "SE", "95% CI low", "95% CI high", "p"))
```

```{r lmm-slopes}
b <- fixef(model_ri)
slopes <- c(Standard = b[["time_numeric"]],
            New      = b[["time_numeric"]] + b[["time_numeric:treatment"]])
round(slopes, 3)
```

**Interpretation**

* Under standard treatment the biomarker falls by about **0.90 units per month**.
* The **time × treatment interaction (−0.56 per month, 95% CI shown above)** means
  the new-treatment group declines faster, by about **1.46 units per month**.
  Over 12 months that adds up to roughly 6.7 extra units of reduction.
* The `treatment` coefficient (−3.8) is the group difference **at baseline**. It
  reflects the baseline imbalance, not an effect of treatment, because no
  treatment has acted at month 0.
* **Age:** each additional year is associated with a slightly higher biomarker
  (≈ 0.14 units). **Sex:** the confidence interval is wide and includes zero, so
  there is no clear sex difference.

## Variance components and ICC

```{r icc}
vc <- as.data.frame(VarCorr(model_ri))
sigma2_b <- vc$vcov[vc$grp == "id"]
sigma2_e <- vc$vcov[vc$grp == "Residual"]
icc <- sigma2_b / (sigma2_b + sigma2_e)

tibble(Component = c("Between-patient (random intercept)", "Within-patient (residual)"),
       Variance  = c(sigma2_b, sigma2_e),
       SD        = sqrt(c(sigma2_b, sigma2_e))) |>
  kable(digits = 2)
round(icc, 3)
```

The **ICC ≈ 0.57**. After accounting for time, treatment, age and sex, 57% of the
remaining variance is due to stable differences between patients. Two
measurements from the same patient therefore correlate at about 0.57. Ignoring
this clustering (ordinary regression) would understate the standard errors of
between-patient effects.

**The random intercept** \(b_{0i}\) is each patient's personal deviation from the
population baseline level. Patients differ in where they start, with an SD of
about `r round(sqrt(sigma2_b), 1)` units, but in this model they all share the same
treatment-specific rate of change.

```{r fig-ranef, fig.height = 3.4}
re <- ranef(model_ri)$id |> rownames_to_column("id") |> rename(b0 = `(Intercept)`)
p_re <- ggplot(re, aes(sample = b0)) +
  stat_qq(alpha = 0.5, size = 1) + stat_qq_line(colour = "#D95F0E") +
  labs(title = "Predicted random intercepts (Q-Q)", x = "Theoretical quantiles", y = "b0")
save_fig(p_re, "03_random_intercepts_qq.png", h = 3.4)
```

# Do patients differ in their rate of change? (random slopes)

```{r lmm-rs}
model_rs <- lmer(
  biomarker ~ time_numeric * treatment + age_c + sex + (time_numeric | id),
  data = data_long, REML = TRUE
)
VarCorr(model_rs)
isSingular(model_rs)
```

The two models have the same fixed effects, so I compare them with a
likelihood-ratio test on the **REML** fits (`refit = FALSE`). The ML refit used by
default is meant for comparing fixed effects.

```{r lrt}
lrt <- anova(model_ri, model_rs, refit = FALSE)
lrt

# The null hypothesis puts the slope variance on the boundary of its space
# (sigma^2_slope = 0), so the naive chi-square(2) p-value is conservative.
# The correct reference is a 50:50 mixture of chi-square(1) and chi-square(2).
chisq <- lrt$Chisq[2]
p_mixture <- 0.5 * pchisq(chisq, 1, lower.tail = FALSE) +
             0.5 * pchisq(chisq, 2, lower.tail = FALSE)
round(c(LRT = chisq, p_naive = lrt$`Pr(>Chisq)`[2], p_mixture = p_mixture), 3)
```

**Which model?** The random-intercept model is preferred.

* *Statistically:* the random-slope variance is essentially zero (a singular,
  boundary fit). The likelihood-ratio test is non-significant even with the
  correct boundary-adjusted mixture p-value, and AIC/BIC do not favour the larger
  model.
* *Practically:* patients differ mainly in **level**, not in **rate** of decline.
  The simpler model is stable, easier to explain, and gives the same fixed
  effects. Its residual plot (below) shows no structure left over.

```{r fig-lmm-resid, fig.height = 3.6}
p_fit <- tibble(fitted = fitted(model_ri), resid = resid(model_ri)) |>
  ggplot(aes(fitted, resid)) +
  geom_hline(yintercept = 0, colour = "grey50") +
  geom_point(alpha = 0.35, size = 1) +
  geom_smooth(se = FALSE, colour = "#D95F0E", method = "loess") +
  labs(title = "Random-intercept model: residuals vs fitted", x = "Fitted", y = "Residual")
save_fig(p_fit, "04_lmm_residuals.png", h = 3.6)
```

# Generalized estimating equations

GEE models the **population-averaged** mean and treats the within-patient
correlation as a nuisance. Robust (sandwich) standard errors keep inference valid
even when the working correlation is misspecified.

```{r gee}
data_gee <- data_long |> arrange(id, time_numeric)

gee_ind  <- geeglm(biomarker ~ time_numeric * treatment, id = id, data = data_gee,
                   family = gaussian, corstr = "independence")
gee_exch <- geeglm(biomarker ~ time_numeric * treatment, id = id, data = data_gee,
                   family = gaussian, corstr = "exchangeable")
gee_ar1  <- geeglm(biomarker ~ time_numeric * treatment, id = id, data = data_gee,
                   family = gaussian, corstr = "ar1")

gee_tab <- map_dfr(
  list(Independence = gee_ind, Exchangeable = gee_exch, `AR(1)` = gee_ar1),
  ~ tidy(.x, conf.int = TRUE), .id = "Working correlation"
) |>
  select(`Working correlation`, term, estimate, std.error, conf.low, conf.high)
kable(gee_tab, digits = 3)

c(exchangeable_alpha = gee_exch$geese$alpha, ar1_alpha = gee_ar1$geese$alpha) |> round(3)
sapply(list(Independence = gee_ind, Exchangeable = gee_exch, `AR(1)` = gee_ar1),
       function(m) QIC(m)[c("QIC", "CIC")]) |> round(1)
```

**Comparing working correlations.** Independence and exchangeable give the same
point estimates. This is expected here: every patient has the same four visit
times and treatment is constant within a patient, so weighting by an
exchangeable correlation does not change the least-squares solution. Their robust SEs are also nearly identical. AR(1)
changes the estimates slightly. QIC mainly reflects the mean model; CIC, the
penalty part, is designed for choosing the working correlation. Neither suggests
AR(1) is the better choice. The main finding (time × treatment ≈ −0.53 to −0.56)
holds under every structure.

## Comparison with the mixed model (like for like)

The task's GEE has no age or sex, while the mixed model has both. For a fair
comparison I also fit the GEE with the **same covariates** as the mixed model.

```{r gee-vs-lmm}
gee_exch_adj <- geeglm(biomarker ~ time_numeric * treatment + age_c + sex, id = id,
                       data = data_gee, family = gaussian, corstr = "exchangeable")

cmp <- full_join(
  tidy(model_ri, effects = "fixed") |> select(term, LMM = estimate, LMM_SE = std.error),
  tidy(gee_exch_adj) |> select(term, GEE = estimate, GEE_robust_SE = std.error),
  by = "term"
)
kable(cmp, digits = 3)
```

With the same covariates, the mixed model and GEE agree almost exactly. **This
is expected, not a coincidence.** For a continuous outcome with an
**identity link**, the subject-specific coefficients of a linear mixed model are
the same as the population-averaged coefficients: averaging
\(\beta_0 + b_{0i} + \dots\) over patients just removes \(b_{0i}\). The two
approaches still answer different questions:

* **GEE:** how does the *average* biomarker in the population change, and how
  does that average differ between treatment groups? Robust to misspecified
  correlation, but makes no statements about individuals and needs data
  missing completely at random.
* **Mixed model:** how does *a given patient's* biomarker change? It models the
  correlation explicitly, gives patient-level predictions, and stays valid when
  data are missing at random.

With a **non-linear link** (e.g. logistic regression for a binary outcome) the
two sets of coefficients differ: subject-specific effects are larger in magnitude
than population-averaged ones. The choice between them is then substantive,
not just technical.

# Survival: Kaplan–Meier and log-rank test

```{r km}
km_fit <- survfit(Surv(follow, death) ~ trt, data = data_wide)
km_fit
```

```{r fig-km, fig.height = 4.6}
km_df <- tidy(km_fit) |>
  mutate(trt = factor(sub("trt=", "", strata), levels = c("Standard", "New")))
km_start <- km_df |> distinct(trt) |>
  mutate(time = 0, estimate = 1, conf.low = 1, conf.high = 1, n.censor = 0)
km_df <- bind_rows(km_start, km_df) |> arrange(trt, time) |>
  group_by(trt) |> mutate(time_next = lead(time, default = max(time))) |> ungroup()

lr <- survdiff(Surv(follow, death) ~ trt, data = data_wide)
lr_p <- pchisq(lr$chisq, df = 1, lower.tail = FALSE)

p_km <- ggplot(km_df, aes(time, estimate, colour = trt, fill = trt)) +
  geom_step(linewidth = 0.9) +
  geom_rect(aes(xmin = time, xmax = time_next,
                ymin = conf.low, ymax = conf.high), alpha = 0.10, colour = NA) +
  geom_point(data = filter(km_df, n.censor > 0), shape = 3, size = 1.6) +
  scale_colour_manual(values = trt_cols) + scale_fill_manual(values = trt_cols) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  annotate("text", x = 5.2, y = 0.95, label = sprintf("Log-rank p = %.3f", lr_p)) +
  labs(x = "Years since baseline", y = "Survival probability",
       colour = "Treatment", fill = "Treatment",
       title = "Kaplan–Meier survival by treatment (95% CI; + = censored)") +
  theme(legend.position = "top")
save_fig(p_km, "05_kaplan_meier.png", h = 4.6)
```

```{r km-table}
summary(km_fit, times = c(1, 2, 3, 5))$table
summary(km_fit, times = c(1, 2, 3, 5)) |>
  (\(s) tibble(Treatment = sub("trt=", "", s$strata), Year = s$time,
               `At risk` = s$n.risk, Survival = s$surv,
               `95% CI` = sprintf("%.2f–%.2f", s$lower, s$upper)))() |>
  kable(digits = 2)
lr
```

**Findings.** The curves separate within the first year, when about half of
all deaths occur, and stay apart. Estimated 3-year survival is 57% under
standard treatment and 73% under the new treatment. Median survival is
3.5 years (95% CI from 2.9) under standard treatment and is not reached under
the new treatment. Survival is consistently higher in the
new-treatment group. Under standard treatment 60 of 134 patients died,
compared with 31 of 106 under the new treatment. The log-rank test rejects equal
survival distributions (χ² = 6.8, 1 df, p = 0.009). The new group had fewer deaths
than expected under the null hypothesis, and the standard group more. Few patients remain at
risk after about five years, so the right-hand ends of the curves are imprecise.

# Survival: Cox proportional hazards model

```{r cox1}
cox1 <- coxph(Surv(follow, death) ~ treatment + age + sex, data = data_wide)
hr_table <- function(m) {
  tidy(m, exponentiate = TRUE, conf.int = TRUE) |>
    transmute(Term = term, HR = estimate, `95% CI low` = conf.low,
              `95% CI high` = conf.high, p = p.value)
}
kable(hr_table(cox1), digits = 3)
```

* **Treatment: HR ≈ 0.56 (95% CI 0.36–0.87).** At any time, a patient on the new
  treatment has an estimated 44% lower instantaneous risk of death than a
  comparable patient (same age and sex) on standard treatment. The CI excludes 1.
* **Age: HR ≈ 1.03 per year**, i.e. about 3% higher hazard per year of age
  (≈ 37% per decade); the lower CI limit is close to 1.
* **Sex:** HR ≈ 1.13 with a wide CI that includes 1, so there is no evidence of a sex difference.

## Proportional hazards assumption

The model assumes each hazard ratio is **constant over follow-up**. The
standard check plots scaled Schoenfeld residuals against time: a flat smooth
supports the assumption, while a trend means the effect changes over time. The
plots are supported by the Grambsch–Therneau test. For continuous covariates,
linearity of the log-hazard (e.g. martingale residuals) and influential
observations are also worth checking.

```{r ph-test}
zph1 <- cox.zph(cox1)
zph1
```

```{r fig-schoenfeld, fig.height = 3.6, fig.width = 10}
zph_df <- map_dfr(colnames(zph1$y), \(v)
  tibble(var = v, time = zph1$x, resid = zph1$y[, v]))
coef_df <- tibble(var = names(coef(cox1)), beta = coef(cox1))
p_sch <- ggplot(zph_df, aes(time, resid)) +
  geom_hline(data = coef_df, aes(yintercept = beta), colour = "grey40", linetype = "dashed") +
  geom_point(alpha = 0.35, size = 0.9) +
  geom_smooth(method = "loess", se = TRUE, colour = "#D95F0E", fill = "#D95F0E", alpha = 0.15) +
  facet_wrap(~ var, scales = "free_y") +
  labs(title = "Scaled Schoenfeld residuals (KM-transformed time)",
       x = "Transformed time", y = "β(t) residual")
save_fig(p_sch, "06_schoenfeld.png", w = 10, h = 3.6)
```

The dashed line is each covariate's estimated log hazard ratio. Under
proportional hazards, the smooth \(\hat\beta(t)\) should stay flat around it.
None of the smooths shows a clear trend (the wiggles stay inside wide bands),
and none of the tests is significant
(global p ≈ 0.67), so proportional hazards is a reasonable assumption here.

# Is the baseline biomarker prognostic?

```{r cox2}
cox2 <- coxph(Surv(follow, death) ~ treatment + age + sex + bio_0, data = data_wide)
kable(hr_table(cox2), digits = 3)

# HR per 10-unit higher baseline biomarker (a more readable scale)
round(exp(10 * c(coef(cox2)["bio_0"], confint(cox2)["bio_0", ])), 2)

AIC(cox1, cox2)
anova(cox1, cox2)   # likelihood-ratio test for adding bio_0
```

```{r compare-trt}
bind_rows(
  `Treatment + age + sex`         = hr_table(cox1) |> filter(Term == "treatment"),
  `Treatment + age + sex + bio_0` = hr_table(cox2) |> filter(Term == "treatment"),
  .id = "Model"
) |> select(-Term) |> kable(digits = 3)
```

**Interpretation**

* Adjusting for the baseline biomarker moves the treatment HR from **0.56 to 0.61**.
  The new treatment remains protective (CI 0.39–0.96).
* `bio_0` is measured **before treatment starts**, so it cannot be a mediator.
  Treatment cannot act *through* a value recorded before it began. The change in
  HR is **confounding adjustment**: the new-treatment group happened to start
  with lower biomarker values (Section 2). If lower baseline values carry lower
  risk, part of the crude survival advantage reflects that head start, and
  adjusting removes it. In a properly randomised trial the adjusted estimate is
  the more precise of the two.
* **The evidence that `bio_0` is prognostic is weak.** Each 10-unit higher baseline
  value is associated with about a 41% higher hazard (95% CI 0.94–2.10). The CI
  includes 1, the likelihood-ratio test gives p ≈ 0.09, and AIC improves by less
  than 1 point. The direction is
  plausible; the precision is limited.

## How longitudinal biomarkers and survival relate

In clinical studies a biomarker can be (i) a **prognostic marker** of disease
severity, (ii) a **mediator** on the causal path from treatment to outcome, or
(iii) a candidate **surrogate endpoint**. Here, the new treatment both lowers the
biomarker faster (Sections 5–7) and improves survival (Sections 8–9). That is
consistent with the biomarker lying on the causal pathway, but these analyses
cannot show it.

Answering that question would require modelling the **post-baseline**
trajectory. Putting later biomarker values directly into a Cox model as fixed
covariates would be wrong: it conditions on the future and creates
*immortal-time bias*. The data also show why caution is needed. All 240
patients have a 12-month biomarker value, yet
`r sum(data_wide$follow < 1 & data_wide$death == 1)` patients died **before**
12 months. That is an artefact of the simulation, but in real data the
biomarker would be missing after death (informative dropout). The appropriate
tools are:

* a **joint longitudinal–survival model** (e.g. `JMbayes2`, `joineRML`), which
  links each patient's current modelled biomarker value or slope to their hazard
  and handles informative dropout; or
* a **landmark analysis**: among patients alive at a landmark time, use the
  biomarker history up to that time to predict survival afterwards.

# Summary of findings

| Question | Method | Result |
|---|---|---|
| Does the biomarker change over time? | RM-ANOVA, LMM | Declines in both groups (≈ 0.90 units/month under standard treatment) |
| Does the change differ by treatment? | RM-ANOVA, LMM, GEE | Faster decline under the new treatment: extra −0.56 units/month (≈ −1.46 vs −0.90) |
| Where does the variation lie? | Variance components | ICC ≈ 0.57; patients differ in level, not in rate (random slope variance ≈ 0) |
| Does the new treatment improve survival? | KM, log-rank, Cox | Yes: log-rank p = 0.009; adjusted HR 0.56 (0.36–0.87); PH holds |
| Is baseline biomarker prognostic? | Cox | Weak evidence (CI includes 1); adjusting for it moves the treatment HR to 0.61 (0.39–0.96) |

# What was revised from the original coursework {#revisions}

1. **Mediation claim corrected.** The original text said the treatment HR shifted
   because part of the effect works "through biomarker levels". Because `bio_0`
   is a pre-treatment variable, the shift is confounding adjustment for a
   baseline imbalance, not mediation.
2. **Sphericity checked.** Mauchly's test and Greenhouse–Geisser-corrected tests
   are now reported (sphericity holds, though strict compound symmetry does not), and residuals come from the repeated-measures model rather
   than from an `lm()` that ignores patients.
3. **Fair GEE-vs-LMM comparison.** The GEE is also fitted with the mixed model's
   covariates. The near-identical estimates are explained by the identity link.
4. **Boundary-corrected LRT.** The random-slope test uses REML fits and the
   χ²₁:χ²₂ mixture reference distribution.
5. **Uncertainty reported.** Standard errors and confidence intervals accompany
   all fixed effects, and age is centred so the intercept is meaningful.
6. **Schoenfeld residual plots** are now shown, not just described.
7. **Reproducibility.** One R Markdown file runs from the raw data to every table
   and figure, with package versions recorded in `renv.lock`.

# Session information

```{r session}
sessionInfo()
```
